One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-23 papers/1_dmatpol/paper-r2.tex
Revision in notation

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i}^{(10)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w}^{(10)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(10)$' denotes the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field and zeroth\hyp{}order with respect to
the electric field gradient.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;10)}_{\alpha\beta} = \Big\{
                               C_{\alpha i} {\bf b}_{\beta i}^{(10)} + C_{\beta i} {\bf b}_{\alpha i}^{(10)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}^{(10)}
%
\Big\}  \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;10)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;10)}$
are unique properties of electronically unperturbed molecule.

@@

It is instructive to start first from the discussion of the polarization under the uniform electric field
condition because, in this case, it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;,
 \end{align}
\end{subequations}
%
and the associated error analysis is shown in Figure~\ref{f:fig-1}.
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{data/dmatpol/water/figure1/fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the density matrix polarization susceptibility tensors of water molecule
in moderate uniform electric field.} The notation $(R_{\max}0)$ denotes the maximum rank of the electric field
and zero rank of electric field gradient in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0 to 0.035 a.u. were tested in 100 random samples with random direction of electric field.
} 
\end{figure}

>>>>

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i}^{(10)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w}^{(10)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(10)$' denotes the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field and zeroth\hyp{}order with respect to
the electric field gradient.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;10)}_{\alpha\beta} = %\Big\{
                               C_{\alpha i} {\bf b}_{\beta i}^{(10)} + C_{\beta i} {\bf b}_{\alpha i}^{(10)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}^{(10)}
%
%\Big\} 
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;10)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;10)}$
are unique properties of electronically unperturbed molecule.

@@

It is instructive to start first from the discussion of polarization under the uniform electric field
condition because, in this case, it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and the associated error analysis is shown in Figure~\ref{f:fig-1}.
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{data/dmatpol/water/figure1/fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the density matrix polarization susceptibility tensors of water molecule
in moderate uniform electric field.} The notation $(R_{\max}0)$ denotes the maximum rank of the electric field
and zero rank of electric field gradient in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0 to 0.035 a.u. were tested in 100 random samples with random direction of electric field.
} 
\end{figure}
%
All the models that are linear with respect to the electric field incorrectly predict the polarization energy.
Interestingly, the predicted values are in perfect correlation with the exact estimates with a slope of $\sim$2.
Incorporation of the second\hyp{}order term from Eq.~\eqref{e:B20} 
obtained by numerical differentiation dramatically reduces the errors
leading to quantitative accuracy (with RMSE of 0.029 kcal/mol and $R^2=99.58\%$;
see also red filled dots in Figure~\ref{f:fig-1}).
What is particularly
